Nonparametric Clustering from Energy Statistics

### 2017-08-13 qcqp_clustering/draft/energy_clustering.tex
new experiments, polished draft

\begin{abstract}
Energy statistics 
was proposed by 
Sz\' ekely in the 80's
inspired by the Newtonian gravitational potential from classical mechanics,
and it provides a nonparametric test for equality of distributions.
It was generalized to 
metric spaces of strong negative type,
and more recently a connection with reproducing kernel Hilbert spaces (RKHS)
was proposed.
Although  
extensively used in statistics, it
has not been previously incorporated in machine learning problems.
Here we consider the problem of clustering data from
an energy statistics theory perspective.
We provide a precise mathematical formulation yielding
a quadratically constrained quadratic program (QCQP), which we show
to be equivalent
to kernel $k$-means optimization problem.
Thus, our results imply a first principles derivation of 
kernel $k$-means from energy statistics.
Moreover, we also consider a weighted version of energy statistics 
applied to clustering, which makes connections to 
graph partitioning problems.
To find local optimizers of such QCQP we consider an iterative
algorithm based on Hartigan's method, which in this case
has the same computational cost 
as kernel $k$-means algorithm but usually with better clustering
performance. We provide carefully designed numerical experiments
showing the superiority of this method compared to
kernel $k$-means, standard $k$-means, and gaussian mixture models.
\end{abstract}

@@

In the experiments below we fix the semimetric 
according to the traditional energy distance \eqref{eq:energy}, and
the point $x_0=0$ is chosen in the corresponding kernel  
\eqref{eq:kernel_semimetric}. We thus have
\begin{equation}
\label{eq:standard_metric}
\rho(x,y) = \| x-y\| \qquad \mbox{and} \qquad \kk(x,y) = 
\tfrac{1}{2}\left( \| x \| + \| y \| - \| x-y \| \right).
\end{equation}
We will consider other kernels as well, 
but \eqref{eq:standard_metric} will be considered the
standard kernel for energy statistics and will always be present in every
experiment for reference. 
%Notice that this is a convention, 
%we could have chosen
%any other semimetric/kernel as the standard.

Let us briefly mention that we compared Algorithm~\ref{algo} and
Algorithm~\ref{algo1d} for several univariate distributions, and
both perform almost indistinguishable regarding the clustering quality.
We omit these results since we will analyse more interesting scenarios
in higher dimensions and with more than two clusters.

The main goal of the 
experiments shown in this section 
is to compare Algorithm~\ref{algo}, based on Hartigan's method, to
kernel $k$-means as described in Algorithm~\ref{kmeans_algo}, based on
Lloyd's method. From the discussion in the previous section we expect 
that Algorithm~\ref{algo}  be superior than Algorithm~\ref{kmeans_algo}.
Another goal is to illustrate the nonparametric aspect of energy statistics.
To this end we also compare
Algorithm~\ref{algo} to standard $k$-means and GMM/EM since 
these are the most used
clustering algorithms in practice.
Moreover, for every algorithm, we always
choose the initialization procedure from $k$-means++%
\footnote{Notice that we just use the initialization
procedure and not the full $k$-means++ algorithm.} \cite{Vassilvitskii}.
Our measure of clustering quality will be the accuracy \eqref{eq:accuracy}
based on the ground truth. Furthermore, in every experiment we sample
data many times and show the average value of the accuracy, with error bars
indicating standard error. Whenever possible, for reference, 
we also show the best possible classification 
accuracy obtained from a numerical estimation
of Bayes classification error, which uses the true probability distribution
of the data.

From the results of \cite{Telgarsky}, the improvement of Hartigan's method
over Lloyd's method should be more prominent in high dimensions.
To analyse this we show 
how the algorithms degrade as the number of dimensions increase, while
keeping the number of points in each cluster fixed. Consider
two clusters with multivariate normal distributions given by
\begin{equation}
\label{eq:gauss1}
\begin{split}
x &\in \C_i  \sim 
\mathcal{N}(\mu_i,\Sigma_i) \qquad (i=1,2),  \\
\mu_1 &= (\underbrace{0,\dotsc,0}_{\times D})^\top , \quad
\mu_2 = 0.7 \times (\underbrace{1,\dots,1}_{\times 10},
\underbrace{0,\dots,0}_{\times (D-10)})^\top, \quad
\Sigma_1 = \Sigma_2 = I_D.
\end{split}
\end{equation}
Note that the Bayes error
is also kept fixed as $D$ increases, given an optimal classification
accuracy of $\approx 0.86$.
We keep signal only the first $10$ dimensions of $\mu_2$ and increase $D$.
For each $D$ we sample $100$ times, picking $100$ points for each cluster.
We apply each algorithm to the resulting dataset 
and compute the average value of the accuracy \eqref{eq:accuracy}.
For both Algorithm~\ref{algo} and Algorithm~\ref{kmeans_algo} 
we use the standard
kernel \eqref{eq:standard_metric} from energy statistics.
The results are shown in Fig.~\ref{fig:gauss}a.
Notice that GMM/EM algorithm is unable to estimate the covariance matrix 
when the number of dimensions exceeds the number of points in each cluster.
We see that Algorithm~\ref{algo} performs better than all the other ones,
and in particular it works better than kernel $k$-means as the number
of dimensions increase. Note also that although GMM and $k$-means are
parametric, they are consistent models for the above distributions.
Now we consider the same experiment but with different covariances,
\begin{equation}
\label{eq:gauss2}
\begin{split}
x &\in \C_i  \sim 
\mathcal{N}(\mu_i,\Sigma_i) \qquad (i=1,2),  \\
\mu_1 &= (\underbrace{0,\dotsc,0}_{\times D})^\top , \quad
\mu_2 = (\underbrace{1,\dots,1}_{\times 10},
\underbrace{0,\dots,0}_{\times (D-10)})^\top, \quad
\Sigma_1 = I_D, \quad 
\Sigma_2 = \left( \begin{smallmatrix} \widetilde{\Sigma} & 0 \\ 0 & I_{D-5}
\end{smallmatrix}\right)
\end{split}
\end{equation}
where $\widetilde{\Sigma}_{jj} = j$ for $j=1,\dotsc,5$.
The best possible accuracy from Bayes classification error is
$\approx 0.95$.
The results are shown in Figure~\ref{fig:gauss}b.

In the experiments of Figure~\ref{fig:gauss}
we can see a better performance of Algorithm~\ref{algo} compared
to the other ones, and in particular to kernel $k$-means,
where we recall that both find approximate solutions
to the same optimization problem \eqref{eq:qcqp2}.
Notice that Algorithm~\ref{algo} is more robust as the number of dimensions
increase.

\begin{figure}
\begin{minipage}{0.49\textwidth}
\centering
\includegraphics[width=1\textwidth]{gauss_dim.pdf}\\[-1.0em]
(a)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\centering
\includegraphics[width=1\textwidth]{gauss_cov.pdf}\\[-1.0em]
(b)
\end{minipage}
\caption{
\label{fig:gauss}
Comparison of Algorithm~\ref{algo}, Algorithm~\ref{kmeans_algo},
standard $k$-means and GMM/EM as the number of dimensions increase.
We sample $100$ times and show the average of \eqref{eq:accuracy} with
error bars being standard error. We have two clusters with $100$ points
each. 
(a) Data distributed as in \eqref{eq:gauss1}. The optimal accuracy
from Bayes error is indicated by the dashed line which is $\approx 0.86$.
(b) Data from \eqref{eq:gauss2}. The optimal accuracy is $\approx 0.95$.
}
\end{figure}

In Figure~\ref{fig:unbalanced} we consider the effect of having 
unbalanced clusters. We generate data as
\begin{equation}
\label{eq:gauss3}
\begin{split}
x &\in \C_i \sim  
\dfrac{n_i}{N} \mathcal{N}(\mu_i,\Sigma_i) \quad (i=1,2), \quad 
\mu_1 = (0,0,0,0)^\top , \quad
\mu_2 = 1.5\times (1,1,0,0)^\top, \\
\Sigma_1 &= I_4, \quad
\Sigma_2 = \left( 
\begin{smallmatrix} 
1/2 & 0 & 0 & 0\\
0 & 1/2 & 0 & 0 \\
0 & 0 & 1 & 0 \\
0 & 0 & 0 & 1 
\end{smallmatrix}\right), \quad
n_1 = N - m, \quad  n_2 = N + m, \quad N=200.
\end{split}
\end{equation}
We increase $m$, that is we make the clusters progressively more unbalanced,
and plot the mean of \eqref{eq:accuracy} with standard errors, where
we sample $100$ times for each $m$.
As expected, GMM
works better than the other algorithms since it is a fuzzy clustering.
We can see that the other methods based on hard assignments degrade similarly,
and  more rapidly than GMM. This indicates that a fuzzy version of
Algorithm~\ref{algo} should compensate for this effect.

\begin{figure}
\includegraphics[width=0.5\textwidth]{gauss_pi.pdf}\vspace{-1em}
\caption{
\label{fig:unbalanced}
Comparison of Algorithm~\ref{algo}, Algorithm~\ref{kmeans_algo}, standard
$k$-means and GMM/EM for data distributed as in \eqref{eq:gauss3}, where
we make the two clusters progressively more unbalanced.
}
\end{figure}

Besides the standard kernel from energy statistics 
\eqref{eq:standard_metric}, associated to \eqref{eq:energy}, 
let us consider two other semimetrics with their respective generating
kernels,
\begin{align}
\rho_{1/2}(x,y) &= \| x-y \|^{1/2} & 
 \kk_{1/2}(x,y) &= \tfrac{1}{2} \left( 
\| x \|^{1/2} + \| y \|^{1/2} 
- \| x-y \|^{1/2} \right), \label{eq:rhohalf}\\
\rho_{e}(x,y) &= 
2 - 2 e^{-\tfrac{1}{2}\| x- y\|} &
 \kk_{e}(x,y) &= e^{-\tfrac{1}{2}\| x-y\|}.
\label{eq:rhoe}
\end{align}
The kernel \eqref{eq:rhohalf} corresponds to the energy distance
\eqref{eq:energy2} with $\alpha=1/2$.
Now consider data normally distributed in $D=20$ as follows:
\begin{equation}
\label{eq:20gauss}
\begin{split}
x &\in \C_i \sim \mathcal{N}(\mu_i,\Sigma_i) \qquad (i=1,2), \\
\mu_1 &= (\underbrace{0,\dotsc,0}_{\times 20})^\top ,\quad
\mu_2 = \tfrac{1}{2} 
(\underbrace{1,\dotsc,1}_{5},\underbrace{0,\dotsc,0}_{15})^\top, \quad
\Sigma_1 = \tfrac{1}{2} I_{20},  \quad
\Sigma_2 = I_{20}.
\end{split}
\end{equation}
We sample an equal number of points for each cluster, which is progressively
increased. The optimal accuracy of \eqref{eq:20gauss} based on Bayes
classification error is $\approx 0.90$. We repeat the same experiment as
before but now using also the kernels \eqref{eq:rhohalf} and
\eqref{eq:rhoe}. The results are shown in Fig.~\ref{fig:consist}a.
Algorithm~\ref{algo} performs better than all the other ones, and in 
particular the kernel \eqref{eq:rhoe} provides better results.
As the number of points get large enough, GMM starts to approach
optimal Bayes accuracy, 
as it should since it is 
consistent model to the distribution of the data. However, 
Algorithm~\ref{algo} with kernel \eqref{eq:rhoe} achieves optimal accuracy
with a much smaller number of points. Moreover, Algorithm~\ref{algo} performs
better than Algorithm~\ref{kmeans_algo} for any of the kernel choices.
Standard $k$-means performs poorly in this example.

\begin{figure}
\begin{minipage}{0.49\textwidth}
\centering
\includegraphics[width=1\textwidth]{gauss.pdf}\\[-1.0em]
(a)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\centering
\includegraphics[width=1\textwidth]{loggauss.pdf}\\[-1.0em]
(b)
\end{minipage}
\caption{
\label{fig:consist}
Algorithm~\ref{algo} and Algorithm~\ref{kmeans_algo} with kernels
\eqref{eq:standard_metric}, \eqref{eq:rhohalf} and \eqref{eq:rhoe}, 
$k$-means and GMM. The optimal accuracy in both cases
is $\approx 0.9$. We show the average of \eqref{eq:accuracy}
over $100$ samples with standard error.
(a) Data distributed as in 
\eqref{eq:20gauss}. 
(b) Data as in \eqref{eq:20loggauss}.
}
\end{figure}

Consider the same parameters as in \eqref{eq:20gauss} but now with
lognormal distributions, 
\begin{equation}
\label{eq:20loggauss}
\log x \in \C_i \sim \mathcal{N}(\mu_i, \Sigma_i) \qquad (i=1,2).
\end{equation}
The same previous experiment is shown in 
Fig.~\ref{fig:consist}b.
Note that Algorithm~\ref{algo} still performs accurately, 
while $k$-means works almost at chance,
and GMM is not even able to estimate the parameters. Again, the
kernel \eqref{eq:rhoe}
provides better results than \eqref{eq:standard_metric} or
\eqref{eq:rhohalf}. Algorithm~\ref{algo} performs better
than Algorithm~\ref{kmeans_algo} for every kernel choice. 
The experiments in Fig.~\ref{fig:consist}
illustrates that energy statistics clustering is nonparametric.

\begin{figure}
\begin{minipage}{0.24\textwidth}
\includegraphics[width=1\textwidth]{2cigars.pdf}\\[-1em] (a)
\end{minipage}
\begin{minipage}{0.24\textwidth}
\includegraphics[width=1\textwidth]{2circles.pdf}\\[-1em] (b)
\end{minipage}
\begin{minipage}{0.24\textwidth}
\includegraphics[width=1\textwidth]{3circles.pdf}\\[-1em] (c)  
\end{minipage}
\begin{minipage}{0.23\textwidth}
\vspace{-0.9em}
\includegraphics[width=1\textwidth]{mnist.pdf}\\[-1.8em] (d)  
\end{minipage}
\caption{\label{fig:other}
(a) Parallel cigars. (b) Two  
concentric circles with noise. (c) Three
concentric circles. (d) MNIST handwritten digits.
Clustering results are in Table~\ref{table:other}
and Table~\ref{table:mnist}.
}
\end{figure}

Now consider the following choices of semimetric and 
corresponding generating kernel:
\begin{align}
\rho_{\alpha}(x,y) &= \| x - y\|^\alpha & 
\kk_{\alpha}(x,y) &= \tfrac{1}{2}\left(
\| x \|^\alpha +
\| y \|^\alpha -
\| x-y \|^\alpha \right) ,
\label{eq:rhoalpha} \\
%
\widetilde{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|}{2 \sigma}} &
\widetilde{\kk}_{\sigma}(x,y) &= e^{-\tfrac{\|x-y\|}{2\sigma}} ,
\label{eq:rhoesigma} \\
%
\widehat{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|^2}{2 \sigma^2}} &
\widehat{\kk}_{\sigma}(x,y) &= e^{-\tfrac{\|x-y\|^2}{2\sigma^2}} .
\label{eq:rhogsigma} 
\end{align}
In Fig.\ref{fig:other} we have examples of
complex two dimensional datasets. For each class in Fig.\ref{fig:other}a
we sample $200$ points. In each class of 
Fig.\ref{fig:other}b and Fig.\ref{fig:other}c we sample $400$ points.
We apply Algorithm~\ref{algo} and Algorithm~\ref{kmeans_algo}, besides
$k$-means and GMM, using the above kernels. The results are shown
in Table~\ref{table:other} where the respective choice of parameters for the
kernels are indicated. In the data shown in Fig.~\ref{fig:other}a 
the standard semimetrics from energy statistics, $\rho_1$ and $\rho_{1/2}$, are
able to provide more accurate results compared to $k$-means. However,
the gaussian kernel $\widetilde{\rho}_2$ gives results very accurate
results, similar to GMM,
which is the true model of the data. In Fig.~\ref{fig:other}b we see
that the choice of kernel is very sensitive, and only \eqref{eq:rhoesigma}
was able to cluster accurately. The same kernel choice to the case
of Fig.~\ref{fig:other} still provides better results than
the other choices, but the results are less accurate.

>>>>

\begin{abstract}
Energy statistics 
was proposed by 
Sz\' ekely in the 80's
inspired by the Newtonian gravitational potential from classical mechanics,
and it provides a nonparametric test for equality of distributions.
It was generalized to 
metric spaces of strong negative type,
and more recently a connection with reproducing kernel Hilbert spaces (RKHS)
was proposed.
Although  
extensively used in statistics, it
has not been previously incorporated in machine learning problems.
Here we consider the problem of clustering data from
an energy statistics theory perspective.
We provide a precise mathematical formulation yielding
a quadratically constrained quadratic program (QCQP), which we show
to be equivalent
to kernel $k$-means optimization problem.
Thus, our results imply a first principles derivation of 
kernel $k$-means from energy statistics.
Moreover, we also consider a weighted version of energy statistics 
applied to clustering, which makes connection to 
graph partitioning problems.
To find local optimizers of such QCQP we consider an iterative
algorithm based on Hartigan's method, which in this case
has the same computational cost 
as kernel $k$-means algorithm but usually with better clustering
performance. We provide carefully designed numerical experiments
showing the superiority of this method compared to
kernel $k$-means, standard $k$-means, and gaussian mixture models.
\end{abstract}

@@

In the experiments below we fix the semimetric 
according to the traditional energy distance \eqref{eq:energy}, and
the point $x_0=0$ is chosen in the corresponding kernel  
\eqref{eq:kernel_semimetric}. Therefore,
\begin{equation}
\label{eq:standard_metric}
\rho(x,y) = \| x-y\| \qquad \mbox{and} \qquad \kk(x,y) = 
\tfrac{1}{2}\left( \| x \| + \| y \| - \| x-y \| \right).
\end{equation}
We will consider other kernels as well 
but \eqref{eq:standard_metric} will be the
standard kernel for energy statistics and will always be present in the
experiments as a reference.

Let us briefly mention that we compared Algorithm~\ref{algo} and
Algorithm~\ref{algo1d} for several univariate distributions and
both perform almost indistinguishable regarding the clustering quality.
We omit these results since we will analyse more interesting scenarios.

The main goal of the 
experiments  to follow
is to compare Algorithm~\ref{algo} based on Hartigan's method, to
kernel $k$-means, as described in Algorithm~\ref{kmeans_algo}, which
is based on Lloyd's method. 
From the discussion in the previous section we expect 
that Algorithm~\ref{algo}  be superior than Algorithm~\ref{kmeans_algo}.
Another goal is to illustrate the nonparametric aspect of energy statistics.
To this end we also compare
Algorithm~\ref{algo} to standard $k$-means and GMM/EM since 
these are the most used
clustering algorithms in practice.
Moreover, for every algorithm, we always
choose the initialization procedure from $k$-means++%
\footnote{Notice that we just use the initialization
procedure and not the full $k$-means++ algorithm.} \cite{Vassilvitskii}.
Our measure of clustering quality will be the accuracy \eqref{eq:accuracy}
based on the ground truth. Furthermore, in every experiment, we sample
data many times and show the average value of the accuracy with error bars
indicating the standard error. Whenever possible, 
we also indicate the optimal 
accuracy computed from
Bayes classification error.

From the results of \cite{Telgarsky}, the improvement of Hartigan's 
over Lloyd's method should be more prominent in high dimensions.
Thus, let us analyse
how the algorithms degrade as the number of dimensions increase, while
keeping the number of points in each cluster fixed. Consider
two clusters with multivariate normal distributions given by
\begin{equation}
\label{eq:gauss1}
\begin{split}
x &\in \C_i  \sim 
\mathcal{N}(\mu_i,\Sigma_i) \qquad (i=1,2),  \\
\mu_1 &= (\underbrace{0,\dotsc,0}_{\times D})^\top , \quad
\mu_2 = 0.7 \times (\underbrace{1,\dots,1}_{\times 10},
\underbrace{0,\dots,0}_{\times (D-10)})^\top, \quad
\Sigma_1 = \Sigma_2 = I_D.
\end{split}
\end{equation}
Note that the Bayes error
is fixed as $D$ increases, giving an optimal 
accuracy of $\approx 0.86$.
For each $D$ we sample $100$ times, picking $100$ points for each cluster.
We apply each algorithm to the resulting dataset 
and compute the average of the accuracy \eqref{eq:accuracy} (error bars
are standard error).
Algorithm~\ref{algo} and Algorithm~\ref{kmeans_algo} 
use the standard
kernel \eqref{eq:standard_metric}.
The results are shown in Fig.~\ref{fig:gauss}a.
GMM/EM is unable to estimate the covariance matrices
when the number of dimensions exceeds the number of points in each cluster.
We see that Algorithm~\ref{algo} performs better than all the other ones,
and in particular it outperforms kernel $k$-means as the number
of dimensions increase. Note also that although GMM and $k$-means are
parametric they are consistent models to the data.

\begin{figure}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=1\textwidth]{gauss_dim.pdf}\\[-1.0em] (a)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=1\textwidth]{gauss_cov_squareroot.pdf}\\[-1.0em] (b)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=1\textwidth]{gauss_cov_linear.pdf}\\[-1.0em] (c)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\includegraphics[width=1\textwidth]{gauss_cov_square.pdf}\\[-1.0em] (d)
\end{minipage}
\caption{
\label{fig:gauss}
Comparison of Algorithm~\ref{algo}, Algorithm~\ref{kmeans_algo},
standard $k$-means and GMM/EM as the number of dimensions increase.
We compute the average of \eqref{eq:accuracy} over $100$ samples with
error bars being standard error. We have two clusters with $100$ points
each. 
(a) Data as in \eqref{eq:gauss1}, where the optimal accuracy
from Bayes error is the dashed line equal to $\approx 0.86$.
(b) Data from \eqref{eq:gauss2} with $q=1/2$ in \eqref{eq:cov}. 
(c) Data from \eqref{eq:gauss2} with $q=1$ in \eqref{eq:cov}.
(d) Data from \eqref{eq:gauss2} with $q=2$ in \eqref{eq:cov}.
The optimal accuracy in these last cases is $1$.
}
\end{figure}

Let us now consider the same experiment as before with data distributed as
\begin{equation}
\label{eq:gauss2}
x \in \C_i  \sim 
\mathcal{N}(\mu_i,\Sigma_i) \quad (i=1,2),  \quad
\mu_1 = (\underbrace{0,\dotsc,0}_{\times D})^\top , \quad
\mu_2 = (\underbrace{1,\dots,1}_{\times 10},
\underbrace{0,\dots,0}_{\times (D-10)})^\top, 
\end{equation}
where
\begin{equation}
\label{eq:cov}
(\Sigma_1)_{ij} = \begin{cases}
i^{-q} \delta_{ij} & \mbox{if $i \le 10$} \\
\delta_{ij} & \mbox{if $10 < i \le D$}
\end{cases} \qquad
(\Sigma_2)_{ij} = \begin{cases}
i^{\,q} \delta_{ij} & \mbox{if $i \le 10$} \\
\delta_{ij} & \mbox{if $10 < i \le D$}
\end{cases} 
\end{equation}
and we choose $q \in \{ 1/2, 1, 2\}$. In these cases,
the best possible accuracy from Bayes classification error is $\approx 1$.
In Fig.~\ref{fig:gauss}b--d we have $q=1/2$, $q=1$, and $q=2$, respectivelly.
GMM/EM was unnable to estimate the covariance matrices
as dimensions get large. Algorithm~\ref{algo} outperforms
Algorithm~\ref{kmeans_algo}, and $k$-means degrades faster as 
$q$ increases.

To summarize, in the experiments of Fig.~\ref{fig:gauss}
we see a better performance of Algorithm~\ref{algo} compared
to the other ones, and in particular to kernel $k$-means,
where we recall that both find approximate and local solutions
to the same optimization problem \eqref{eq:qcqp2}.
Algorithm~\ref{algo} is more robust as the number of dimensions
increase.

Let us now
consider the effect of having 
unbalanced clusters. We generate data as
\begin{equation}
\label{eq:gauss3}
\begin{split}
x &\in \C_i \sim  
\dfrac{n_i}{N} \mathcal{N}(\mu_i,\Sigma_i) \quad (i=1,2), \quad 
\mu_1 = (0,0,0,0)^\top , \quad
\mu_2 = 1.5\times (1,1,0,0)^\top, \\
\Sigma_1 &= I_4, \quad
\Sigma_2 = \left( 
\begin{smallmatrix} 
1/2 & 0 & 0 & 0\\
0 & 1/2 & 0 & 0 \\
0 & 0 & 1 & 0 \\
0 & 0 & 0 & 1 
\end{smallmatrix}\right), \quad
n_1 = N - m, \quad  n_2 = N + m, \quad N=200.
\end{split}
\end{equation}
We then increase $m$, 
that is we make the clusters progressively more unbalanced,
and plot the average of \eqref{eq:accuracy} 
over
$100$ samples for each $m$
(error bars are standard error). The results are in Fig.~\ref{fig:unbalanced}.
As expected, GMM
works better than the other algorithms since it is a fuzzy clustering.
The other methods based on hard assignments degrade similarly
and  more rapidly than GMM. This indicates that a fuzzy version of
Algorithm~\ref{algo} should compensate for this effect.

\begin{figure}
\includegraphics[width=0.5\textwidth]{gauss_pi.pdf}\vspace{-1em}
\caption{
\label{fig:unbalanced}
Comparison of Algorithm~\ref{algo}, Algorithm~\ref{kmeans_algo}, 
$k$-means, and GMM/EM. The data is distributed as \eqref{eq:gauss3} where
we make the clusters progressively more unbalanced.
}
\end{figure}

Besides the standard kernel from energy statistics 
\eqref{eq:standard_metric}, 
let us consider two other semimetrics with their respective generating
kernels,
\begin{align}
\rho_{1/2}(x,y) &= \| x-y \|^{1/2} & 
 \kk_{1/2}(x,y) &= \tfrac{1}{2} \left( 
\| x \|^{1/2} + \| y \|^{1/2} 
- \| x-y \|^{1/2} \right), \label{eq:rhohalf}\\
\rho_{e}(x,y) &= 
2 - 2 e^{-\tfrac{1}{2}\| x- y\|} &
 \kk_{e}(x,y) &= e^{-\tfrac{1}{2}\| x-y\|}.
\label{eq:rhoe}
\end{align}
The kernel \eqref{eq:rhohalf} corresponds to the energy distance
\eqref{eq:energy2} with $\alpha=1/2$.
Now consider data normally distributed in $D=20$ as
\begin{equation}
\label{eq:20gauss}
\begin{split}
x &\in \C_i \sim \mathcal{N}(\mu_i,\Sigma_i) \qquad (i=1,2), \\
\mu_1 &= (\underbrace{0,\dotsc,0}_{\times 20})^\top ,\quad
\mu_2 = \tfrac{1}{2} 
(\underbrace{1,\dotsc,1}_{5},\underbrace{0,\dotsc,0}_{15})^\top, \quad
\Sigma_1 = \tfrac{1}{2} I_{20},  \quad
\Sigma_2 = I_{20}.
\end{split}
\end{equation}
We sample an equal number of points for each cluster, which is progressively
increased. The optimal accuracy based on Bayes
classification error is $\approx 0.90$. 
Clustering results are shown in Fig.~\ref{fig:consist}a.
Algorithm~\ref{algo} outperforms the other ones, and in 
particular the kernel \eqref{eq:rhoe} provides better results.
As the number of points get large enough GMM starts to approach
optimal Bayes, as it should since it is a
consistent model to the data. However, 
Algorithm~\ref{algo} with kernel \eqref{eq:rhoe} approach optimal Bayes
with a much smaller number of points. Moreover, Algorithm~\ref{algo} 
outperforms
Algorithm~\ref{kmeans_algo} for any of the kernel choices.
%Standard $k$-means performs poorly in this example.

Now consider the same parameters as in \eqref{eq:20gauss} but with
lognormal distributions, 
\begin{equation}
\label{eq:20loggauss}
\log x \in \C_i \sim \mathcal{N}(\mu_i, \Sigma_i) \qquad (i=1,2).
\end{equation}
The same previous experiment is shown in 
Fig.~\ref{fig:consist}b.
Note that Algorithm~\ref{algo} still performs accurately, 
while $k$-means works almost at chance,
and GMM is not even able to estimate the parameters. Again, the
kernel \eqref{eq:rhoe}
provides better results than \eqref{eq:standard_metric} or
\eqref{eq:rhohalf}. 
The experiments in Fig.~\ref{fig:consist}
illustrate how energy statistics clustering is nonparametric.

\begin{figure}
\begin{minipage}{0.49\textwidth}
\centering
\includegraphics[width=1\textwidth]{gauss.pdf}\\[-1.0em]
(a)
\end{minipage}
\begin{minipage}{0.49\textwidth}
\centering
\includegraphics[width=1\textwidth]{loggauss.pdf}\\[-1.0em]
(b)
\end{minipage}
\caption{
\label{fig:consist}
Algorithm~\ref{algo} and Algorithm~\ref{kmeans_algo} with kernels
\eqref{eq:standard_metric}, \eqref{eq:rhohalf} and \eqref{eq:rhoe}, 
$k$-means, and GMM. The optimal accuracy in both cases
is $\approx 0.9$. We show the average of \eqref{eq:accuracy}
over $100$ samples with standard error.
(a) Data distributed as in 
\eqref{eq:20gauss}. 
(b) Data distributed as in \eqref{eq:20loggauss}.
}
\end{figure}

\begin{figure}
\begin{minipage}{0.24\textwidth}
\includegraphics[width=1\textwidth]{2cigars.pdf}\\[-1em] (a)
\end{minipage}
\begin{minipage}{0.24\textwidth}
\includegraphics[width=1\textwidth]{2circles.pdf}\\[-1em] (b)
\end{minipage}
\begin{minipage}{0.24\textwidth}
\includegraphics[width=1\textwidth]{3circles.pdf}\\[-1em] (c)  
\end{minipage}
\begin{minipage}{0.23\textwidth}
\vspace{-0.9em}
\includegraphics[width=1\textwidth]{mnist.pdf}\\[-1.8em] (d)  
\end{minipage}
\caption{\label{fig:other}
(a) Parallel cigars. (b) Two  
concentric circles with noise. (c) Three
concentric circles with noise. (d) MNIST handwritten digits.
Clustering results are in Table~\ref{table:other}
and Table~\ref{table:mnist}.
}
\end{figure}

Consider the following choices of semimetric and 
corresponding generating kernel:
\begin{align}
\rho_{\alpha}(x,y) &= \| x - y\|^\alpha & 
\kk_{\alpha}(x,y) &= \tfrac{1}{2}\left(
\| x \|^\alpha +
\| y \|^\alpha -
\| x-y \|^\alpha \right) ,
\label{eq:rhoalpha} \\
%
\widetilde{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|}{2 \sigma}} &
\widetilde{\kk}_{\sigma}(x,y) &= e^{-\tfrac{\|x-y\|}{2\sigma}} ,
\label{eq:rhoesigma} \\
%
\widehat{\rho}_{\sigma}(x,y) &= 2 - 2 e^{-\tfrac{\|x-y\|^2}{2 \sigma^2}} &
\widehat{\kk}_{\sigma}(x,y) &= e^{-\tfrac{\|x-y\|^2}{2\sigma^2}} .
\label{eq:rhogsigma} 
\end{align}
In Fig.\ref{fig:other} we have examples of
complex two dimensional datasets. The two parallel cigars of 
Fig.~\ref{fig:other}a have $200$ points each. For the concentric circles
of Fig.~\ref{fig:other}b and Fig.~\ref{fig:other}c we
sample $400$ points for each class.
We apply Algorithm~\ref{algo} and Algorithm~\ref{kmeans_algo}, besides
$k$-means and GMM, using the above kernels. The results are shown
in Table~\ref{table:other} where the respective choice of parameters for the
kernels are indicated. For the data in Fig.~\ref{fig:other}a 
the semimetrics $\rho_1$ and $\rho_{1/2}$ are
able to provide more accurate results compared to $k$-means. However,
the gaussian kernel $\widetilde{\rho}_2$ gives very accurate
results, similar to GMM,
which is the true model of the data. For the data shown in
Fig.~\ref{fig:other}b we see
that the choice of kernel is very sensitive, and only \eqref{eq:rhoesigma}
was able to cluster accurately. The same kernel choice to the case
of Fig.~\ref{fig:other} still provides better results than
the others, but they are less accurate compared to Fig.~\ref{fig:other}b.

\newcolumntype{g}{>{\columncolor{gray!20}}l}
\begin{table}[h]
\renewcommand*{\arraystretch}{0.75}
\begin{tabular}{@{}r  l g  l g  l g@{}}
\toprule[1pt]
 & & \emph{Fig.~\ref{fig:other}a}
 & & \emph{Fig.~\ref{fig:other}b}
 & & \emph{Fig.~\ref{fig:other}c} \\
\midrule[0.5pt]
\multirow{4}{*}{\emph{Algorithm \ref{algo}}~~~~} 
& $\rho_{1}$ & $0.766\pm0.066$
& $\rho_{1}$ & $0.522\pm0.006$
& $\rho_{1}$ & $0.437\pm0.030$ \\
& $\rho_{1/2}$ & $0.859\pm0.062$
& $\rho_{1/2}$ & $0.524\pm0.007$
& $\rho_{1/2}$ & $0.547\pm0.026$ \\
& $\widetilde{\rho}_{2}$ & $0.971\pm0.015$
& $\widetilde{\rho}_{1}$ & $\bm{0.9999\pm0.0001}$
& $\widetilde{\rho}_{2}$ & $\bm{0.677\pm0.003}$ \\
& $\widehat{\rho}_{2}$  & $\bm{0.998\pm0.001}$
& $\widehat{\rho}_{1}$  & $0.597\pm0.052$
& $\widehat{\rho}_{2}$ & $0.645\pm0.012$ \\
\arrayrulecolor{gray!80}\midrule[0.5pt]
\multirow{4}{*}{\emph{Algorithm \ref{kmeans_algo}}~~~~} 
& $\rho_{1}$ & $0.758\pm 0.069$
& $\rho_{1}$ & $0.516\pm 0.002$
& $\rho_{1}$  & $0.452\pm 0.030$ \\
& $\rho_{1/2}$ & $0.901\pm 0.060$
& $\rho_{1/2}$ & $0.524\pm 0.007$
& $\rho_{1/2}$ & $0.570\pm 0.016$ \\
& $\widetilde{\rho}_{2}$ & $0.971\pm 0.015$
& $\widetilde{\rho}_{1}$ & $\bm{0.9999\pm 0.0001}$
& $\widetilde{\rho}_{2}$ & $\bm{0.673\pm 0.002}$ \\
& $\widehat{\rho}_{2}$ & $\bm{0.998\pm 0.001}$
& $\widehat{\rho}_{1}$ & $0.528\pm 0.008$
& $\widehat{\rho}_{2}$ & $0.640\pm 0.013$ \\
\arrayrulecolor{gray!80}\midrule[0.5pt]
\emph{$k$-means}~~~~ 
& & $0.599\pm0.046$
& & $0.521\pm0.005$
& & $0.360\pm0.004$ \\
\emph{GMM}~~~~
& & $0.9995\pm0.0003$
& & $0.598\pm0.018$
& & $0.479\pm0.021$ \\
\arrayrulecolor{black}\bottomrule[1pt]
\end{tabular}
\caption{\label{table:other}
Clustering the data shown in Fig.~\ref{fig:other} with Algorithm~\ref{algo}
and Algorithm~\ref{kmeans_algo}, besides $k$-means and GMM, with kernels 
\eqref{eq:rhoalpha}--\eqref{eq:rhogsigma}. 
We sample $10$ times and show the average
accuracy \eqref{eq:accuracy} with standard error.
}
\end{table}
